Simple repression draft

### 2020-10-17 doc/resubmission/appendix_03_bayesian_inference.tex
send Poisson promoter inference to appendix

Having introduced the ideas behind Bayesian inference we are ready to apply the
theoretical machinery to our non-equilibrium models. In particular in this
section we will focus on model 1 and model 5 in
Figure~\ref{fig2:constit_cartoons}(A). Model 1, the Poisson promoter, will help
us build practical intuition into the implementation of the Bayesian inference
pipeline as we noted in Section~\ref{sec:beyond_means} of the main text that
this model cannot be reconciled with experimental data from observables such as
the Fano factor. In other words, we acknowledge that this model is ``wrong,''
but we still see value in going through the analysis since the simple nature of
the model translates into a neat statistical analysis.

@@

As specified in the main test, the mRNA steady-state distribution for model 1 in
Figure~\ref{fig2:constit_cartoons}(A) is Poisson with parameter $\lambda$.
Throughout this Appendix we will appeal to the convenient notation for
probability distributions of the form
\begin{equation}
m \sim \text{Poisson}(\lambda),
\end{equation}
where the simbol ``$\sim$'' can be read as \textit{is distributed according to}.
So the previous equation can be read as: the mRNA copy number $m$ is distributed
according to a Poisson distribution with parameter $\lambda$. Our objective then
is to compute the posterior probability distribution $P(\lambda \mid D)$, where,
as in the main text, $D = \{ m_1, m_2, \ldots, m_N \}$ are the data consisting
of single-cell mRNA counts. Since we can assume that each of the cells mRNA
counts are independent of any other cells, our likelihood function $P(D \mid
\lambda)$ consists of the product of $N$ Poisson distributions.

>>>>

Having introduced the ideas behind Bayesian inference we are ready to apply the
theoretical machinery to our non-equilibrium models. In particular in this
section we will focus on model 1 and model 5 in
Figure~\ref{fig2:constit_cartoons}(A). Model 1, the Poisson promoter, will help
us build practical intuition into the implementation of the Bayesian inference
pipeline. As we noted in Section~\ref{sec:beyond_means} of the main text, this
model cannot be reconciled with experimental data from observables such as the
Fano factor. In other words, we acknowledge that this model is ``wrong,'' but we
still see value in going through the analysis since the simple nature of the
model translates into a neat statistical analysis.

@@

Model 1 in Figure~\ref{fig2:constit_cartoons}(A) predicts a  mRNA distribution
whose steady-state solution is given by a Poisson distribution with parameter
$\lambda \equiv r / \gamma$, where $r$ is the mRNA production rate, and $\gamma$
is the mRNA degradation rate~\cite{Sanchez2013}. The goal of our inference
problem is then to find the probability distribution for the parameter value
$\lambda$ given the experimental data. By Bayes' theorem this can be written as
\begin{equation}
p(\lambda \mid D) = {p(D \mid \lambda) p(\lambda) \over p(D)},
\end{equation}
where $D = \{m_1, m_2, \ldots, m_N \}$ are the single-cell mRNA experimental
counts. As is standard we will neglect the denominator $p(D)$ on the right
hand side since it is independent of $\lambda$ and serves only as a
normalization factor.

The steady-state solution for the master equation defines the likelihood term
for a single cell $p(m \mid \lambda)$. What this means is that for a given
choice of parameter $\lambda$, under model 1 of
Figure~\ref{fig2:constit_cartoons}(A), we expect to observe $m$ mRNAs in a
single cell with probability
\begin{equation}
p(m\mid\lambda) = \frac{\lambda^m e^{-\lambda}}{m!}.
\label{eq:poisson_inference010}
\end{equation}
Assuming each cell's mRNA count in our dataset is independent of others, the
likelihood of the full inference problem $p(D\mid\lambda)$ is simply a product
of the single cell likelihoods given by Eq.~\ref{eq:poisson_inference010} above, so
\begin{equation}
p(D\mid\lambda) = \prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}.
\end{equation}

Throughout this Appendix we will appeal to the convenient notation for
probability distributions of the form
\begin{equation}
m \sim \text{Poisson}(\lambda),
\end{equation}
where the simbol ``$\sim$'' can be read as \textit{is distributed according to}.
So the previous equation can be read as: the mRNA copy number $m$ is distributed
according to a Poisson distribution with parameter $\lambda$. Our objective then
is to compute the posterior probability distribution $P(\lambda \mid D)$, where,
as in the main text, $D = \{ m_1, m_2, \ldots, m_N \}$ are the data consisting
of single-cell mRNA counts. Since we can assume that each of the cells mRNA
counts are independent of any other cells, our likelihood function $P(D \mid
\lambda)$ consists of the product of $N$ Poisson distributions.

### 2020-10-17 doc/resubmission/section_04_bayesian_inference.tex
minimal changes to fix some errors
send Poisson promoter inference to appendix

Our objective for this section will then be to assess whether or not model 5 is
quantitatively able to reproduce experimental data. In other words, if our claim
is that the level of coarse graining in this model is capable of capturing the
relevant features of the data, then we should be able to find values for the
model parameters that can match theoretical predictions with single-molecule
mRNA count distributions. A natural language for this parameter inference
problem is that of Bayesian probability. We will then build a Bayesian inference
pipeline to fit the model parameters to data. To gain intuition on how this
analysis is done we will begin with the ``wrong'' model 1 in
Figure~\ref{fig2:constit_cartoons}(A). We will use the full dataset of
single-cell mRNA counts from~\cite{Jones2014} used in
Figure~\ref{fig2:constit_cartoons}(B).

\subsubsection{Model 1: Poisson promoter}

Model 1 in Figure~\ref{fig2:constit_cartoons}(A) predicts a  mRNA distribution
whose steady-state solution is given by a Poisson distribution with parameter
$\lambda \equiv r / \gamma$, where $r$ is the mRNA production rate, and $\gamma$
is the mRNA degradation rate~\cite{Sanchez2013}. The goal of our inference
problem is then to find the probability distribution for the parameter value
$\lambda$ given the experimental data. By Bayes' theorem this can be written as
\begin{equation}
p(\lambda \mid D) = {p(D \mid \lambda) p(\lambda) \over p(D)},
\end{equation}
where $D = \{m_1, m_2, \ldots, m_N \}$ are the single-cell mRNA experimental
counts. As is standard we will neglect the denominator $p(D)$ on the right
hand side since it is independent of $\lambda$ and serves only as a
normalization factor.

The steady-state solution for the master equation defines the likelihood term
for a single cell $p(m \mid \lambda)$. What this means is that for a given
choice of parameter $\lambda$, under model 1 of
Figure~\ref{fig2:constit_cartoons}(A), we expect to observe $m$ mRNAs in a
single cell with probability
\begin{equation}
p(m\mid\lambda) = \frac{\lambda^m e^{-\lambda}}{m!}.
\label{eq:poisson_inference010}
\end{equation}
Assuming each cell's mRNA count in our dataset is independent of others, the
likelihood of the full inference problem $p(D\mid\lambda)$ is simply a product
of the single cell likelihoods given by Eq.~\ref{eq:poisson_inference010} above, so
\begin{equation}
p(D\mid\lambda) = \prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}.
\end{equation}

To proceed we need to specify a prior distribution $p(\lambda)$. In this case we
are extremely data-rich, as the dataset from Jones et.\ al~\cite{Jones2014} has
of order 1000-3000 single-cell measurements for each promoter, so our choice of
prior matters little here, as long as it is sufficiently broad. For details on
the prior selection we refer the reader to Appendix~\ref{sec:bayesian}. For our
purpose here it suffices to specify that we use as prior a Gamma distribution.
This particular choice of prior introduces two new parameters, $\alpha$ and
$\beta$, which parametrize the gamma distribution itself, which we use to encode
the range of $\lambda$ values we view as reasonable. Recall $\lambda$ is the
mean steady-state mRNA count per cell, which \textit{a priori} could plausibly
be anywhere from 0 to a few hundred. $\alpha=1$ and $\beta=1/50$ achieve this,
since the gamma distribution is strictly positive with mean $\alpha/\beta$ and
standard deviation $\sqrt{\alpha}/\beta$.

As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior is
known as the \textit{conjugate} prior for a Poisson likelihood.
Conjugate priors have the convenient properties that a closed form exists for the posterior distribution $p(\lambda \mid D)$ - unusual in Bayesian inference problems - and the closed form posterior takes the same form as the prior. For
our case of a Poisson distribution likelihood with its
Gamma distribution conjugate prior, the posterior distribution is also a Gamma
distribution~\cite{Gelman2013}. Specifically the two parameters $\alpha'$ and
$\beta'$ for this posterior distribution take the form $\alpha' = \alpha +
\bar{m} N$ and $\beta' = \beta + N$, where we defined the sample mean $\bar{m} =
\frac{1}{N}\sum_{k=1}^N m_k$ for notational convenience, and $N$ is the number
of cells in our dataset. Furthermore, given that $N$ is $\mathcal{O}(10^3)$ and
$\langle m\rangle \gtrsim 0.1$ for all promoters measured in~\cite{Jones2014}
our data easily overwhelms the choice of prior, and allows us to approximate the
Gamma distribution with a Gaussian distribution with mean $\bar{m}$ and variance
$\bar{m} / N$ with marginal errors. As an example with real numbers, for the
\textit{lacUV5} promoter, Jones et.\ al~\cite{Jones2014} measured 2648 cells
with an average mRNA count per cell of $\bar{m} \approx 18.7$. For this case our
posterior distribution $P(\lambda \mid D)$ would be a Gaussian distribution with
mean $\mu = 18.7$, and a standard deviation $\sigma \approx 0.08$. This suggests
we have inferred our model's one parameter to a precision of order 1\%.

We remind the reader that we began this section claiming that the Poisson model
was ``wrong'' since it could not reproduce features of the data such as a Fano
factor $> 1$. The fact that we obtain such a narrow posterior distribution for our
parameter $P(\lambda \mid D)$ does not equate to the model being adequate to
describe the data. What this means is that given the data $D$, only values
in a narrow range are remotely plausible for the parameter $\lambda$,
but a narrow posterior distribution does not necessarily
mean the model accurately depicts reality.
As we will see later in Figure~\ref{fig:constit_post_full} after
exploring the bursty promoter model, indeed the correspondence
when contrasting the Poisson model with the experimental data is quite poor.

@@

We carried out Markov-Chain Monte Carlo (MCMC) sampling on the posterior of this
model, starting with the constitutive \textit{lacUV5} dataset
from~\cite{Jones2014}. The resulting MCMC samples are shown in
Figure~\ref{fig:constit_post_full}(A). In contrast to the active/inactive
constitutive model considered in~\cite{Razo-Mejia2020} (nonequilibrium model 4
in Figure~\ref{fig2:constit_cartoons}(A)), this model is well-identified with both
parameters determined to a fractional uncertainty of 5-10\%. The strong
correlation reflects the fact that their product sets the mean of the mRNA
distribution, which is tightly constrained by the data, but there is weak
``sloppiness''~\cite{Transtrum2015} along a set of values with a similar
product.

Having found the model's posterior to be well-identified as with the Poisson
promoter, the next step is to compare both models with experimental data. To do
this for the case of the bursty promoter, for each of the parameter samples
shown in Figure~\ref{fig:constit_post_full}(A) we generated negative
bionomial-distributed mRNA counts. As MCMC samples parameter space
proportionally to the posterior distribution, this set of random samples span
the range of possible values that we would expect given the correspondence
between our theoretical model and the experimental data. A similar procedure can
be applied to the Poisson promoter. To compare so many samples with the actual
observed data, we can use empirical cumulative distribution functions (ECDF) of
the distribution quantiles. This representation is shown in
Figure~\ref{fig:constit_post_full}(B). In this example, the median for each
possible mRNA count for the Poisson distribution is shown as a dark green line,
while the lighter green contains 95\% of the randomly generated samples. This
way of representing the fit of the model to the data gives us a sense of the
range of data we might consider plausible, under the assumption that the model
is true. For this case, as we expected given our premise of the Poisson promoter
being wrong, it is quite obvious that the observed data, plotted in black is not
consistent with the Poisson promoter model. An equivalent plot for the bursty
promoter model is shown in blue. Again the darker tone shows the median, while
the lighter color encompasses 95\% of the randomly generated samples. Unlike the
Poisson promoter model, the experimental ECDF closely tracks the posterior
predictive ECDF, indicating this model is actually able to generate the observed
data and increasing our confidence that this model
is sufficient to parametrize the physical reality of the system.

@@

Having stated the problem, Bayes' theorem reads
\begin{equation}
p(\vect{k}, k_i, b \mid D)
\propto
p(D \mid\vect{k}, k_i, b) p(\vect{k}, k_i, b),
\end{equation}
where $D$ is again the set of all $N$ observed single-cell mRNA counts
across the various conditions. We assume that individual single-cell
measurements are independent so that the likelihood factorizes as
\begin{equation}
p(D \mid\vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid \vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid k_j^+, k_j^-, k_i, b)
\end{equation}
where $k_j^\pm$ represent the appropriate binding and unbinding
rates out of $\vect{k}$ for the $j$-th measured cell. The probability
$p(m\mid k_j^+, k_j^-, k_i, b)$ appearing in the last expression
is exactly Eq.~\ref{eq:p_m_bursty+rep_appdx}, the steady-state
distribution for our bursty model with repression derived in
Section~\ref{sec:gen_fcn_appdx}, which for completeness we reproduce here as
\begin{equation}
\begin{split}
p(m \mid k_R^+, k_R^-, k_i, b) = & ~\frac{
        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
        }
        {
        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
        }
\frac{b^m}{m!}
\\
&\times {_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b).
\end{split}
\label{eq:p_m_bursty+rep}
\end{equation}
where $_2F_1$ is the confluent hypergeometric function of the second kind and
$\alpha$ and $\beta$, defined for notational convenience, are
\begin{align}
\begin{split}
\alpha &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
\\
\beta &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
\end{split}
\end{align}

>>>>

Our objective for this section will then be to assess whether or not model 5 is
quantitatively able to reproduce experimental data. In other words, if our claim
is that the level of coarse graining in this model is capable of capturing the
relevant features of the data, then we should be able to find values for the
model parameters that can match theoretical predictions with single-molecule
mRNA count distributions. A natural language for this parameter inference
problem is that of Bayesian probability. We will then build a Bayesian inference
pipeline to fit the model parameters to data. We will use the full dataset of
single-cell mRNA counts from~\cite{Jones2014} used in
Figure~\ref{fig2:constit_cartoons}(B). To gain intuition on how this analysis is
done we analyzed the ``wrong'' model 1 in Figure~\ref{fig2:constit_cartoons}(A).
For this analysis, we direct the reader to Appendix~\ref{sec:bayesian} as an
intuitive introduction to the Bayesian framework with a simpler model.

@@

We carried out Markov-Chain Monte Carlo (MCMC) sampling on the posterior of this
model, starting with the constitutive \textit{lacUV5} dataset
from~\cite{Jones2014}. The resulting MCMC samples are shown in
Figure~\ref{fig:constit_post_full}(A). In contrast to the active/inactive
constitutive model considered in~\cite{Razo-Mejia2020} (kinetic model 4 in
Figure~\ref{fig2:constit_cartoons}(A)), this model is well-identified with both
parameters determined to a fractional uncertainty of 5-10\%. The strong
correlation reflects the fact that their product sets the mean of the mRNA
distribution, which is tightly constrained by the data, but there is weak
``sloppiness''~\cite{Transtrum2015} along a set of values with a similar
product.

Having found the model's posterior to be well-identified as with the Poisson
promoter (Appendix~\ref{sec:bayesian}), the next step is to compare both models
with experimental data. To do this for the case of the bursty promoter, for each
of the parameter samples shown in Figure~\ref{fig:constit_post_full}(A) we
generated negative bionomial-distributed mRNA counts. As MCMC samples parameter
space proportionally to the posterior distribution, this set of random samples
span the range of possible values that we would expect given the correspondence
between our theoretical model and the experimental data. A similar procedure can
be applied to the Poisson promoter. To compare so many samples with the actual
observed data, we can use empirical cumulative distribution functions (ECDF) of
the distribution quantiles. This representation is shown in
Figure~\ref{fig:constit_post_full}(B). In this example, the median for each
possible mRNA count for the Poisson distribution is shown as a dark green line,
while the lighter green contains 95\% of the randomly generated samples. This
way of representing the fit of the model to the data gives us a sense of the
range of data we might consider plausible, under the assumption that the model
is true. For this case, as we expected given our premise of the Poisson promoter
being wrong, it is quite obvious that the observed data, plotted in black is not
consistent with the Poisson promoter model. An equivalent plot for the bursty
promoter model is shown in blue. Again the darker tone shows the median, while
the lighter color encompasses 95\% of the randomly generated samples. Unlike the
Poisson promoter model, the experimental ECDF closely tracks the posterior
predictive ECDF, indicating this model is actually able to generate the observed
data and increasing our confidence that this model is sufficient to parametrize
the physical reality of the system.

@@

Having stated the problem, Bayes' theorem reads
\begin{equation}
p(\vect{k}, k_i, b \mid D)
\propto
p(D \mid\vect{k}, k_i, b) p(\vect{k}, k_i, b),
\end{equation}
where $D$ is again the set of all $N$ observed single-cell mRNA counts
across the various conditions. We assume that individual single-cell
measurements are independent so that the likelihood factorizes as
\begin{equation}
p(D \mid\vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid \vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid k_j^+, k_j^-, k_i, b)
\end{equation}
where $k_j^\pm$ represent the appropriate binding and unbinding
rates out of $\vect{k}$ for the $j$-th measured cell. The probability
$p(m\mid k_j^+, k_j^-, k_i, b)$ appearing in the last expression
is exactly Eq.~\ref{eq:p_m_bursty+rep_appdx}, the steady-state
distribution for our bursty model with repression derived in
Section~\ref{sec:gen_fcn_appdx}, which for completeness we reproduce here as
\begin{equation}
\begin{split}
p(m \mid k_R^+, k_R^-, k_i, b) = & ~\frac{
        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
        }
        {
        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
        }
\frac{b^m}{m!}
\\
&\times {_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b).
\end{split}
\label{eq:p_m_bursty+rep}
\end{equation}
where $_2F_1$ is the confluent hypergeometric function of the second kind and
$\alpha$ and $\beta$, defined for notational convenience, are
\begin{equation}
\begin{split}
\alpha &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
\\
\beta &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
\end{split}
\end{equation}
